Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
up to extensibility

\note{Sil}{is any normalization used, for example, to the maximum values found in the analysis?}

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

>>>>

\note{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}

@@

\note{Sil}{i was wondering whether it would be more clear to present the metrics like this:
\item \texttt{I$_1$}: total number of components, and

now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface
	}
Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Number of specific components
involved in workflow executions (\texttt{R$_1$}). The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Number of specific interactions
involved in workflow executions (\texttt{R$_2$}).
\note{Sil}{add here an example of an interaction that is not, for example, data transfers?}
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine \texttt{E$_1$}: number of interactions or
  components that need to be modified to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
  \note{Si}{new interface = new interaction? maybe better to use one term only here. Maybe also refer to how this shows on figure 2}
\item Engine version upgrade \texttt{E$_2$}: number of interactions or
  components modified to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
  \note{Silv}{it seems to me that this metric cannot be calculated in a straighforward manner from the diagrams.. is that correct? if it can, give an example of how this is seen in the diagram}
\item New workflow \texttt{E$_3$}: number of interactions required
to integrate a new workflow in the architecture. Adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions, assuming
  that the engine type and version required to execute this workflow
  are already available.
  \note{sil}{again here it would be good to indicate how this is seen in the diagram}
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}
